Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-12-27 manuscript/referee/reply.tex
reply added

\documentclass[a4paper,prb,twocolumn,longbibliography]{revtex4-1}  %a4paper,prb,twocolumn
%\documentclass[preprint]{revtex4-1}

@@

\title{Jahn-Teller effect with rigid octahedral rotations in perovskites} %Cooperative

%\title{Orbital ordering in perovskites transition metal oxides with rigid octahedral rotations}

\author{Rukhshanda Naheed}
\affiliation{Department of Physics, Quaid-i-Azam University, Islamabad 45320, Pakistan}

\author{M. Ahsan Zeb}
\affiliation{Department of Physics, Quaid-i-Azam University, Islamabad 45320, Pakistan}
\email{ahsan.zeb@hotmail.com}
%\orcidlink{https://orcid.org/0000-0002-3448-2357}

\author{Kashif Sabeeh}
\affiliation{Department of Physics, Quaid-i-Azam University, Islamabad 45320, Pakistan}

\date{\today}
\begin{abstract}

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
anion octahedra around the transition metal ions
 in ABO$_3$ perovskites
 from the viewpoint of Jahn-Teller problem
 and study the
effect of the crystal field of the nearest A and B ion cages 
on the single electron spectrum of the 
central 
transition metal B ions.
While the crystal field of the octahedra,
which
% quenches the orbital angular momentum of the central B ion and 
creates degenerate manifolds $e_g$ and $t_{2g}$,
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field
in a way that
leads to sizeable splittings in 
%the $e_g$ and $t_{2g}$ manifolds of the central B ion.
these manifolds.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the 
%local orbital ordering 
order of the orbitals
are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

\end{abstract}
\maketitle

Perovskites  
exhibit a rich variety of intriguing 
electronic,
magnetic
 and optical
properties~\cite{ZubkoARCMP11,
HwangNM12, 
BhattacharyaARMR14,
HellmanRMP17,
ChenJPCM17,DagottoS05,DagottoMRSB08}
due to the interplay between
spin, orbital, and structural degrees of 
freedom
and widely varying electronic 
correlations.
These materials 
are important
both for practical applications in all sorts of modern devices 
and 
fundamental understanding of various physical 
phenomena~\cite{SalamonRMP01,
WangS03,
DawberRMP05,
SchlomARMR07,
KosterRMP12}.

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can 
distort~\cite{
BersukerCCR75,
KugelSPU82,
MillisN98,
TerakuraPMS07,
HalcrowRSC13,
LufasoAC04}
or simply 
rotate~\cite{
GlazerAC72, ThomasAC96,
WoodwardAC97a,WoodwardAC97b,
HowardAC98,StokesAC02,AngelPRL05}
 to lower the symmetry
and produce, e.g., a rhombohedral, tetragonal or orthorhombic structure.
% cubic perovskite [spacegroup 221] can transform to:
% crystal system: spacegroups
% triclinic: 2
% monoclinic: 11,12,15
% orthorhombic: 59,62,63,71,74
% tetragonal 127,139,140
% rhombohedral/trigonal: 167
% cubic: 204, 221 
Octahedral rotations also occur in other perovskite related structures
and
have been extensively 
studied
due to their crucial role in determining various physical properties,
e.g., in 
ferroelectric~\cite{GhosezNM11,
RondinelliAM12,
SchickPRB14,
SimPRB14,
LeePRL17,
WangPRL22,
ZhouPRB22,
GaoPRB23} and 
multiferroic~\cite{KhomskiiPhys09,
BenedekPRL11,
YangPRL14,
YePRB18,
ZhangPRL20,
ZhangYajunPRL20, 
ZhouPRB21,
ZhangPRL22}
systems.
The interplay of octahedral rotations
and 
spin-orbit coupling~\cite{AmatNL14,KrachJPCM23},
breathing distortions~\cite{BalachandranPRB13},
strain~\cite{Aguado-PuentePRL11},
defects~\cite{JiaPRB22}, and
pressure~\cite{XiangPRB17},
and 
their role in
magnetostructural phase transitions~\cite{JohnsonPRL20},
orbital order~\cite{OgawaPRL12,ChenPRL19}, orbital anisotropy~\cite{HuangPRB21}, 
and
%surface electronic structure under in-plane uniaxial compression~\cite{MoralesPRL23}
surface electronic structure~\cite{MoralesPRL23},
have also been investigated.
% JT interacting with octa tilting:
Since, usually, the octahedra are not just %(Jahn-Teller) 
distorted or rotated but a combination of both,
the influence of these two 
structural mutations %instabilities 
on each other has recently been explored by many authors~\cite{LufasoAC04,CarpenterAC09,ChenPRL19}
as well.

While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap,
KhomskiiCR21}
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained empirically in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor~\cite{Goldschmidt1926,KieslichCS15}. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-bonds-labelled-2}
\includegraphics[width=1\linewidth]{structures-combined-bonds-labelled-2}
\caption{
Crystal structures and coordination cages.
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:struc}
\end{center}
\end{figure}

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL10,CammarataJCP14,LeeCM16,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
~\cite{HerzbergZPC33,BersukerPL66,KristoffelPSSB67,PearsonJMS83,FultonJCP04,BersukerCR13}
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL10}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskite oxides~\cite{CammarataJCP14}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{LeeCM16}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

We take a completely different approach to this problem.
We go back to the basics
and see 
what is it that lifts 
 the orbital degeneracy in the classical JT effect?
It is a specific 
deformation of the crystal field 
%(${\sim z^2}$)
at the transition metal (JT ion) site.
Does the same happen in the case of octahedral tilting and rotations?
While the potential of an octahedral anion cage 
in its own frame
does not change 
%in case of 
with
rigid rotations for obvious reason,
the total crystal field 
does deform %indeed
because
 the rest of the lattice is not only rotated
  in a given octahedron's frame (that turns its non-spherical potential)
 but also distorted owing to the accompanying structural transformation.
%So, in principle, there should be a (first order) JT effect at play.
Using this insight
and elements of crystal field theory~\cite{BetheAP29,VanVleckPR32,PavariniChap},
%here in this manuscript,
we 
consider next two coordination cages around an octahedron
% explore this problem in detail
and find that,
just like a first order JT effect,
 the deformed crystal field 
in this case also lifts the orbital degeneracy,
by a significant amount for typical system parameters,
signalling the instability against the rotations.

% cooperative: pairs of alpha beta....
% terminnology: rotation vs tilt
% coop/consistent rotations/tilting then makes Pbnm etc...

Since the octahedra are connected to each other 
(making a 3-dimensional network in the perovskite structure),
rotating one octahedron rotates all other in the crystal in a consistent fashion~\cite{GlazerAC72};
 see Figs.~\ref{fig:struc}(a,b).
We consider an octahedral \emph{rotation} $\beta$ around $001$ direction
(that transforms the structure from cubic Pm$\bar 3$m to tetragonal P4/mbm)
%({No.~$127$})
followed by a \emph{tilting} $\alpha$ around the new $110$ direction
(transforming to orthorhombic Pbnm, 
%({No.~$62$})
or Imma 
%({No.~$74$})
 at ${\beta=0}$).
We use the rotation and tilt terminology for these 
two angles throughout the rest of the manuscript.

Beyond the anion octahedron,
each transition metal ion B in a cubic perovskite
is coordinated by two cation cages,
a cube of A and an octahedron of B, as shown in Figs.~\ref{fig:struc}(c,d).
Including the 
%electrostatic 
potential of these two ion cages
into the 
crystal field Hamiltonian of 
the central ion B,
we explore the effect 
of the rotation and tilting on 
%the splittings in 
its $t_{2g}$ and $e_g$ manifolds
that are
created primarily by the anion's octahedral potential.
We find that the degeneracy of $t_{2g}$ and $e_g$
is lifted. 
The size of the splitting 
and 
%local orbital ordering
the order of the orbitals
depend
on the relative charges on the two types of cations
and a competition between rotation and tilting.
At typical parameter values,
the splitting is large enough 
to play a major role in 
the corresponding structural instability. 
The qualitative picture developed by our calculations 
holds even when the potential of the full lattice is considered.

% 3 dim hilbert space of t2g, 2 dim hipbert space of eg
% the mixing of t2g states due to rot and tilting is equiv to an artifact in cubic phase
% so it's not similar to SOCJT states at diff energy mix due to distortion.
% exlain it in the results/discussion section.....???

\com{
$\alpha=0, \beta \neq 0$\\
Tetragonal:
127. P 4 / m b m
$\alpha \neq 0,\beta=0$\\
Orthorhombic:
74. I m m a
$\alpha,\beta \neq 0$\\
Orthorhombic:
62. Pbnm
}

The paper is organised as follows.
Section~\ref{sec:formalistm}
presents the formalism and 
Sec.~\ref{sec:JT} reviews the Jahn-Teller effect to
introduce the notation and concepts involved
by considering tetragonal distortion of an octahedron.
%\az{[.... multiple subsections?]}
The model and calculations are presented
in Sec.~\ref{sec:calc},
where
we first discuss the relative contributions of the three closest ion cages
in the cubic structure
 to
the crystal field at the central transition metal site 
and crystal field splitting between its $e_g$ and $t_{2g}$ manifolds
(Sec.~\ref{sec:relativeVoctABO}),
and then 
describe the deformation effects of 
rigid octahedral rotations %(Sec.~\ref{sec:rotation})
on
the structure (Sec.~\ref{sec:structure}) 
and crystal field (Sec.~\ref{sec:deformCF}).
The results are in Sec.~\ref{sec:results},
where we %seperately
present the splitting in the single particle spectrum
belonging to the $t_{2g}$ (Sec.~\ref{sec:t2g})
and $e_g$ (Sec.~\ref{sec:eg-gen}) manifolds,
and finally describe the effect of the full lattice (Sec.~\ref{sec:FL}).
%\az{appendix?}

\section{Formalism: Crystal field Hamiltonian using multipole expansion of the crystal field potential}
\label{sec:formalistm}

Let us introduce the basic formalism we use throughout this manuscript.
The spectrum of an atom or ion
changes in the presence of nearby charged species.
This is described by the crystal field Hamiltonian of the atom
that contains the coupling between its atomic orbitals and/or their energy shifts
induced by the electrostatic potential of those charges.
Since the angular part of atomic orbitals are described by 
the spherical harmonics,
an expansion of the potential at the atom's site 
in the same bases simplifies the analysis.
This is because
the effect of various multipole components of the potential
is governed by certain selection rules 
(enforced by the symmetries of the spherical harmonics
through Gaunt coefficients, see below).

If we have an arrangement of ions with charge $q$ around a given site,
we can express their %(crystal field) 
potential in terms of spherical harmonics~\cite{PavariniChap}, 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{lm}~\left(\frac{r}{a}\right)^l \ylm{lm},
\end{align}
where $a$ sets the length scale, $\hat r=\vec r/r$ is the direction of $\vec r$,
and $V_{l,m}$ are multipole components of the potential
 (for the given site as origin and a chosen orientation).
%and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we have a transition metal atom/ion
at the central site
with
only $d$-orbitals ($l=2$),
the matrix elements of its crystal field Hamiltonian
are then given by,
\begin{align}
H_{m' m''} &= -\frac{q}{a}\sum_{l,m} V_{lm} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2m'}| Y_{lm}| Y_{2m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$
is the expectation of $(r/a)^l$ evaluated in the $d$-orbitals bases
assuming the same radial dependence for all orbitals~\cite{PaxtonNotes}.
The sum is restricted to even values of $l\leq 4$ as 
$C_{l}^{m'm''m}=0$ otherwise,
with $-l \leq m \leq l$.

We consider transition metal oxides as an example perovskite system throughout the rest of the manuscript,
but our calculations and results are general and also apply to perovskite transition metal compounds with other anions, e.g., halides and chalcogenides.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six oxygen (O) ions each with a charge $-q_o$ making an octahedron of size $a$ (where these anions are placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the $d$-orbitals contains 
 only two $l=4$ components, 
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$),
 and is called octahedral field. 
It is given by~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds: 
% $\Delta$ apart:
$H_{mm} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{mm} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

\com{
$V_{oct}(\vec r)$ causes the d-orbitals to split into two manifolds,
$t_{2g}$  and $e_g$, where $t_{2g}$ contains three degenerate states ($xy,yz,zx$) while $e_g$ has two degenerate states ($x^2-y^2,z^2$).
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.
\az{We will see how these degeneracies are lifted as we elongate (or compress)
the octahedron along a ....??}
}

\section{Jahn-Teller Effect}
\label{sec:JT}

The Jahn--Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap,
KhomskiiCR21}
 describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion, as described below.

%\subsection{Dependence of electronic and elastic energies on the distortion}

The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider an 
elongation or a compression of an octahedron along one of its axes by $a\delta$,
where $a$ is the size of the octahedron and lattice constant of the cubic perovskite structure.
The corresponding change in the unit cell volume would be linear in $\delta$
as $\sim a^3 \delta$
so the elastic energy
would change as $\sim\delta^2$.
Assuming a linear change 
 in the electronic spectrum,
  the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are positive constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In general, as long as the two competing terms 
are not the same degree monomial in the distortion,
the ground state can have a finite distortion.
We use this argument to focus only on the changes
 in the electronic energy throughout this paper.

\com{
The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider a 
hypothetical 
distortion of an octahedron 
along two of its axes, say, by $a\delta_1,a\delta_2$,
where $a$ is the lattice constant of the cubic structure.
The corresponding change in the unit cell volume would be linear in $\delta_{1,2}$
as $\sim a^3\left(\delta_1+\delta_1\right)$
so the elastic energy %(for an isotropic system) 
changes as $\sim\left(\delta_1+\delta_1\right)^2$.
A linear change ($\propto \delta_{1/2}$) in the electronic spectrum is required to produce
 a finite distortion in the optimum structure (or the ground state of full electron-lattice/nuclear problem).
}

It is worth reminding that, in the Jahn-Teller problem,
we usually consider single particle spectrum, i.e.,
single electron energy states,
to analyse the ground state of a many electron system,
which is the lowest energy ``electronic configuration'' 
(a combination of occupied single particle states 
out of all available single particle states, 
or a slater determinant of that combination to be more precise!).
So a degenerate set of states that are completely occupied
makes a non-degenerate many electron state because 
only a single configuration/combination is possible.
For a given number of electrons in the d-shell of the transition metal ion,
we can determine the (most likely) configuration(s) or ground state(s)
using Hund's rule and see if it is degenerate or otherwise.

\com{
Instability:\\
 Low/high spin configurations, rules/arguments using single particle spectrum and a given low/high spin state with a given number of electrons:\\
elastic energy vs electronic... number of electrons... multi-electron configurations.... degenerate vs non-degenerate cases...
}

In the following, we 
consider an octahedral environment around a transition metal ion 
and 
illustrate how the changes in the electronic energy 
can be calculated
 when the octahedron is elongated.

\subsection{Tetragonal distortion of an octahedron}
\label{sec:JTtertagonal}

Let us consider elongation of an octahedron % in Sec.~\ref{sec:egt2g} 
along one of its O-B-O bonds. 
Align the cartesian axes along O-B-O bonds and stretch the two bonds along $z$-axis equally by an amount $a\delta$.
%where $a$ is the size of the octahedron (the distance between the O-O along any cartesian axes).
The potential $V^{JT}(\vec r)$ at the central B site (JT ion) of this deformed octahedron 
still contains only a few non-zero components (see appendix for a method to calculate $V_{lm}^{JT}$).
Not only the ratio 
$V_{4,0}^{JT}:V_{4,4}^{JT}$ now deviates from
 ideal 
octahedral field in \eref{eq:voct},
there is an $l=2$ component induced due to the deformation,
\begin{align}
%V_{0,0}^{JT}=& \frac{8 \sqrt{\pi } (2 \delta +3)}{\delta +1},\\
V_{2,0}^{JT}=& -\frac{32 \sqrt{\frac{\pi }{5}} \delta  \left(\delta ^2+3 \delta +3\right)}{(\delta +1)^3},\\
V_{4,0}^{JT}=& \frac{32}{3} \sqrt{\pi } \left(\frac{4}{(\delta +1)^5}+3\right),\\
V_{4,4}^{JT}=& \frac{32 \sqrt{35 \pi }}{3}.
\end{align}

\com{
At $\delta=0$, keeping only the non-spherical components, 
we obtain the
 octahedral potential for a symmetric octahedron~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
\lim_{\delta\to0}V^{JT}(\vec r) \to V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right],
\end{align}
}

The crystal field Hamiltonian of the JT ion,
${H_{m' m''}^{JT} = (q_o/a)\sum_{l,m} V_{lm}^{JT} C_{l}^{m'm''m} D_{l}}$,
still turns out to be diagonal 
(i.e., the deformation $\delta$
will not couple these orbitals to each other),
but the degeneracies of the two manifolds is lifted
and their respective average energies also shifted.
Defining 
$F_n(\delta) = 1-1/(1+\delta)^n$,
we see that
the $t_{2g}$ now splits into two levels,
with energies
\begin{align}
E_{xy} =&
\frac{32}{7}\frac{q_o}{a} \left(
 D_2F_3 - \frac{20}{9}D_4F_5 
\right),\\
E_{yz} =& E_{zx} = -\frac{1}{2} E_{xy},
\end{align}
as measured from their average 
$E_0^{t_{2g}} = -64q_oD_4 \left(3-F_5\right)/9a$.
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&
\frac{32}{7}\frac{q_o}{a} \left(D_2F_3+\frac{5}{3} D_4F_5\right),\\
E_{z^2}=&-E_{x^2-y^2},\\
\end{align}
as measured from their average 
$E_0^{e_{g}} = 32 q_oD_4 \left(3-F_5\right)/3a$.
The crystal field splitting becomes
$\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=
160q_oD_4\left(3-F_5\right)/9a$.
At small $\delta$, 
expanding 
$F_3,F_5$
 up to first order in $\delta$,
we obtain
\begin{align}
\label{eq:JTt2g}
(E_{xy},E_{yz},E_{zx})&= (2, -1, -1)
%\left[\frac{16}{63} (27 \text{D2}-100 \text{D4})\right]
\left[D_2-\frac{100}{27}  D_4\right]\frac{48}{7}\frac{q_o}{a}\delta,\\
\label{eq:JTeg}
(E_{x^2-y^2},E_{z^2})&=(1,-1)
\left[D_2+\frac{25}{9} D_4\right]\frac{96}{7}\frac{q_o}{a}\delta,
\end{align}
whereas
%${\Delta_{\delta}= \frac{48}{7}\frac{q_o}{a}\frac{\delta}{a}}$.
$E_0^{t_{2g}}$, $E_0^{e_{g}}$ and $\Delta$ all are rescaled by a factor of ${1-5\delta/3}$.

\com{
The crystal field splitting becomes
\begin{align}
E_0^{t_{2g}},E_0^{e_{g}},\Delta \to (1-\frac{5}{3}\delta)E_0^{t_{2g}}
E_0^{e_{g}} \to (1-\frac{5}{3}\delta) E_0^{e_{g}}
\Delta \to (1-\frac{5}{3}\delta)\Delta
\end{align}
}

%\subsection{Parameters $D_2$ and $D_4$ of the JT ion/atom}
The terms in square brackets on the right of Eqs.~\ref{eq:JTt2g},\ref{eq:JTeg}
contain $D_2$ and $D_4$. 
$D_4$ is proportional to the crystal field splitting $\Delta$ 
($D_4=\frac{3a}{160q_o} \Delta$) and 
we can estimate its value from realistic values of $\Delta$ for a given octahedron of size $a$ and ion charge $-q_o$. 
Assuming $a=7.0$ bohr and $q_o=2e$ ($e$ being the elementary charge),
$D_4$ ranges from $1 \times 10^{-3}$ to  $7 \times 10^{-3}$
for $\Delta \simeq 0.5-3$~eV.
Typically, ${D_{2}\approx \sqrt{D_{4}}}$,
so $D_{2}/D_{4} \approx 1/\sqrt{D_{4}}$
and it roughly ranges between $29$ and $12$ for $D_4$ in the above range.
This means that, for realistic systems, 
we should always have $D_{2}/D_{4} > 100/27, 25/9 \approx 4$
so 
$D_{2}$ terms in 
Eqs.~\ref{eq:JTt2g},\ref{eq:JTeg}
dominate and determine the sign and size of the splittings.

So far, we have calculated the changes in the single particle energy levels of the B ion. 
Whether the deformation would be energetically favourable 
or not will depend on the number of electrons occupying these levels,
as illustrated below.

%\subsection{Examples}
Let us consider %the following two cases.
%\begin{enumerate}
%\item
a JT/B ion with two electrons.
In the symmetric octahedron case, 
we will have a degenerate (many electron) ground state 
with both electron in the same spin state (Hund's rule) in the $t_{2g}$ manifold.
Distorting the structure $\delta>0$, we lower the ground state energy because 
we can now choose $yz,zx$ orbitals that have lower energy than the original $t_{2g}$ energy of the undistorted structure.
%\item
In case of a JT/B ion with three electrons, the situation is very different.
Assuming Hund's coupling is enough to still produce complete spin polarisation,
there is only one possible orbital configuration, one electron in each of $xy,yz,zx$ orbitals.
This non-degenerate (many electron) ground state
does not lower its energy if the structure is distorted 
because the gain due to $yz,zx$ is already balanced by the loss due to picking the higher energy $xy$ state.
%\end{enumerate}

%\subsection{cooperative JT and orbital order}
%\az{just a brief description... cooperative JT and orbital order}
To complete the discussion of the JT instability,
it is worth mentioning that
JT effect occurs in all kinds of systems---molecules, clusters and crystals.
In crystals, e.g., perovskites, it becomes a cooperative effect where the
elongation of one octahedron is accompanied by a compression of adjacent octahedron,
which minimizes the changes in the unit cell volume and hence the elastic energy cost of the transformation.
Such alternate elongations and compressions also lead to 
an interesting long range orbital ordering where alternate orbitals
are populated on the adjacent JT sites.

\com{
\begin{align}
E_{xy} =&
\frac{32}{7} \left[
\left(1-\frac{1}{(\delta +1)^3}\right) D_2
-
\left(1-\frac{1}{(\delta +1)^5}\right) \frac{20}{9}D_4
\right]\frac{q_o}{a},\\
E_{yz} =& E_{zx} = -3/2 E_{xy},
\end{align}
as measured from their average 
%\begin{align}
$E_0^{t_{2g}} = -64q_oD_4 \left(2+(\delta +1)^{-5}\right)/9a$.
%\frac{64}{9} \left(-\frac{1}{(\delta +1)^5}-2\right) \text{D4}
%\end{align}
Similarly, $e_g$ splits up as
\begin{align}
E_{x^2-y^2}=&\frac{32}{7} \left[\left(1-\frac{1}{(\delta +1)^3}\right) \text{D2}+\frac{5}{3}\left(1-\frac{1}{(\delta +1)^5}\right) \text{D4}\right]\frac{q_o}{a},\\
E_{z^2}=&-E_{x^2-y^2},
\end{align}
measured from their average 
%\begin{align}
$E_0^{e_{g}} = 32 q_oD_4 \left(2+(\delta +1)^{-5}\right)/3a$.
%\frac{32}{3} \left(\frac{1}{(\delta +1)^5}+2\right) \text{D4}
%\end{align}
The crystal field splitting becomes
\begin{align}
\Delta \equiv E_0^{e_{g}}-E_0^{t_{2g}}=\frac{160}{9} \left(\frac{1}{(\delta +1)^5}+2\right) D_4.
\end{align}
}

\section{Model and calculations of ABO$_3$ Perovskites}%}
\label{sec:calc}

We now turn to our model and calculations of the splittings in the 
$t_{2g}$ and $e_{g}$ manifolds due to the deformations 
in the crystal field that are produced as the octahedra rotate or tilt.
First, let us consider the cubic perovskite and see 
how much A and B cages contribute to the octahedral crystal field and associated splitting $\Delta$ between $t_{2g}$ and $e_{g}$ manifolds.

\subsection{Contribution of different coordination cages in a cubic perovskite}
\label{sec:relativeVoctABO}

>>>>

\documentclass[a4paper,prb]{revtex4-1}  %a4paper,prb,twocolumn
%\documentclass[preprint]{revtex4-1}

@@

\textbf{BK14571/Naheed: Jahn-Teller effect with rigid octahedral rotations in perovskites}

\rev{
The paper "Jahn-Teller effect with rigid octahedral rotations in
perovskites" by R. Naheed and co-authors represents in my opinion a
very nice piece of work, where the authors thoroughly study a rather
complicated subject of influence of rotation and tilting on lifting
orbital degeneracy. However, it has to be mentioned that there is one
rather general point, which must be cleared out before the decision
can be made.

The present treatment follows a general route often taken in the JT
physics: to consider the effect of Madelung field, but there is also
another extremely important contribution to the crystal field coming
from the hybridization, i.e. the overlap between orbitals centered on
different sites. Typically it acts in the same direction, but this
effect should be taken into account for the quantitative results, on
which the authors aimed. The angle dependence of the hopping
integrals, can be considered via the Slater-Koster approximation.

Other, more technical comments:
}

\begin{enumerate}
\item
\rev{It is not clearly explained how exactly the authors define sign of
q0, qA,qB. E.g. the charge neutrality relation mentioned above (16) is
$qB = 3q0 - qA$. For oxides q0 is -2 and $qA>0$ and does not typically
exceed +4, but then $qB<0$. Obviously the authors used some other
conventions, but do not really explain this.
}

Thanks for pointing it out.
We take the charge on the anion/oxygen to be $-q_o$ (so $q_B=3q_o-q_A$ is not incorrect).
This is stated at the start of  
Sec.~I-A: Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds (page 3, first column).
For oxides, $-q_o=-2$ (so $q_o=2$), and the charge netrality correctly gives $q_B=6-q_A$;
$q_A, q_B >0$.

In the revised version,
for clarity,
we have added a comment ``(remember that we take anion charge to be $-q_o$)''
just after the charge neutrality relation above Eq.~(16).

\item
\rev{
I'm not sure that it is clear for a general reader why from volume
dependence $V = a^3 (\delta_1 + \delta_2)$ follows that the elastic energy
must be proportional to $(\delta_1 + \delta_2)^2$ [BTW one of the
subscripts is wrong]. OK, they know from the high-school courses that
the elastic term must quadratic and could agree, but why the same
logic applies to $(\alpha^2 + \beta^2)^2$ can be not clear. Moreover,
it's not clear a priori why there is no cubic terms. Please explain
this part in more detail.
}

Thanks. 
The the main point we
tried to make here is that 
the electronic and elastic energies have different dependence on the distortion (so the optimum structure is distorted).
For our purpose, it is not important to know the exact form of the elastic energy and whether anharmonic effects would be important or not.

To avoid any distraction and confusion,
we have rewritten the whole paragraph. 
We now focus
on conveying the key point without
unnecessary details.

\item
\rev{
Fig. 7. Should not be there explained what a bold line means? And
why there is no a bold line for dashed curves?
}

Thanks for noticing. 
What appears to be a bold line are actually two nearly degenerate lines of the same thickness as the other lines in the plot.
In comparison to the dashed lines (nearest AB cages only), 
the Ewald sum
in case of infinite lattice considers
a large number of \emph{unit cells}
that include some \emph{incomplete}
coordination cages at the boundary, 
 which breaks some symmetries present in the 
 case of complete coordination cages.

 To avoid any confusion,
we have included in the caption:
 ``(The bold black line is nothing but an overlap of two practically degenerate black lines.)''.

\item
\rev{
I would also suggest to put labels of the A and B cages to Fig. 1.
}

Thanks. We have done it in the revised version.

\end{enumerate}
